A 1% Increase in Spending on Facilities Decreases Graduation Rates by 0.2%: An Analysis on Ontario School Districts and Related Predictors

### 2024-12-04 paper/paper.qmd
updated miscellaneous aspects of the project like sketches and README

---
title: "Facilities Spending and Socioeconomic Factors Show Strong Correlation with Ontario School District Graduation Rates"
author: Andrew Goh
thanks: "Code and data are available at: https://github.com/mushroomcarbon/schools_and_money"
date: today
date-format: long
abstract: "Using data from Ontario's school board achievement, financial, and demographic datasets, we analyze the relationship between various district-level factors and four-year graduation rates. Through polynomial regression analysis, we find that the percentage of households without post-secondary degrees and facilities spending have significant negative correlations with graduation rates, while per-student expenses show a non-linear relationship. These findings suggest that socioeconomic factors and resource allocation decisions significantly impact academic outcomes, with implications for education policy and resource distribution in Ontario school districts."
format: pdf
header-includes:
  - \usepackage{float}
  - \usepackage{placeins}
number-sections: true
bibliography: references.bib
---

@@

# Load the polynomial model from the RDS file
poly_model <- readRDS(here("models", "poly_model.rds"))
lm_model <- readRDS(here("models", "lm_model.rds"))

@@

The relationship between educational resources and academic outcomes has been a central focus of education policy research, particularly in publicly funded school systems. In Ontario, where school boards receive varying levels of funding and serve communities with diverse socioeconomic profiles, understanding how different factors influence student success is crucial for effective policy-making [@ontarioeducation].

Recent studies have suggested that both financial resources and community characteristics play significant roles in determining student outcomes [@educationresources; @socioeconomic]. However, the specific mechanisms through which these factors interact, particularly in the Ontario context, remain unclear. This gap in understanding is particularly relevant given Ontario's significant investment in education, with annual spending exceeding $31 billion [@ontariobudget].

Our paper examines how various district-level factors—including financial allocations, socioeconomic indicators, and operational characteristics—correlate with four-year graduation rates across Ontario school boards. We focus specifically on the relationships between graduation rates and six key variables: total expenses, percentage of households without post-secondary degrees, percentage of low-income households, percentage of budget spent on facilities, per-student expenses, and total enrollment.

@@

The results suggest that socioeconomic factors, particularly the percentage of households without post-secondary degrees, have a significant negative linear relationship with graduation rates. Additionally, we find that the relationship between per-student expenses and graduation rates follows a U-shaped curve, suggesting diminishing returns up to a certain point. These findings have important implications for how educational resources might be most effectively allocated to improve student outcomes.

@@

## Overview
We use the statistical programming language R [@citer] for the analysis and presentation of the project. The tidyverse [@tidyverse] ecosystem, particularly dplyr [@dplyr], was used for data manipulation. For model development and evaluation, we utilized caret [@caret] for model training and glmnet [@glmnet1; @glmnet2; @glmnet3] for regularized regression methods. The arrow [@arrow] package was employed for efficient data reading and writing, while broom [@broom] was used for converting statistical objects into tidy data frames. Additional file management was facilitated by the here [@here] package.

@@

@tbl-summary presents an overview of the summary statistics of our key variables across Ontario school boards, and @tbl-snapshot presents a snapshot of the data itself.

@@

summary_stats %>%
  mutate(
    across(contains("Total Expenses"), ~./1e9),  # Convert to billions
    across(contains("Total Enrolment"), ~./1000)  # Convert to thousands
  ) %>%
  pivot_longer(everything(), 
               names_to = c("Variable", "Statistic"), 
               names_pattern = "(.+)_(.+)") %>%
  pivot_wider(names_from = Statistic, 
             values_from = value) %>%
  gt() %>%
  fmt_number(
    columns = c(Mean, SD, Min, Max, Median),
    decimals = 2
  ) %>%
  cols_label(
    Variable = "Variable",
    Mean = "Mean",
    SD = "Std. Dev.",
    Min = "Minimum",
    Max = "Maximum",
    Median = "Median"
  ) %>%
  tab_footnote(
    footnote = "Total Expenses shown in billions ($), Enrollment in thousands, percentages as raw values, expenses per student in dollars.",
    locations = cells_column_labels(columns = Mean)
  )
```

@@

```{r}
#| label: tbl-snapshot
#| tbl-cap: "Sample of First Five Rows from Ontario School Board Dataset"
#| echo: false
#| warning: false
#| message: false

# First half of columns
knitr::kable(
  analysis_data |>
    select(`Board Number`,
           `Board Name`,
           Region,
           City,
           `Four Year Graduation Rate`,
           `Total Expenses`) |>
    slice(1:5) |>
    mutate(Region = str_replace(Region, "Region ", "")),  # Remove the word "Region"
  col.names = c("Board ID",
                "Name",
                "Region",
                "City",
                "Grad Rate",
                "Total Expenses ($)"),
  digits = c(0, 0, 0, 0, 3, 3),
  booktabs = TRUE
)

# Second half of columns
knitr::kable(
  analysis_data |>
    select(`Expenditure on Facilities`,
           `Total Enrolment`,
           percent_low_income,
           percent_no_degree,
           percentage_spent_on_facilities,
           expenses_per_quota) |>
    slice(1:5),
  col.names = c("Facility Expenses ($)",
                "Enrollment",
                "Low Income (%)",
                "No Degree (%)",
                "Facilities (%)",
                "$/Student"),
  digits = c(3, 0, 3, 3, 3, 3),
  booktabs = TRUE
)
```

@@

The percentage of students from low-income households ranges from 10.35% to 25.92%, with a mean of 15.77%. This variable provides insight into the socioeconomic composition of each school board's student population.

@@

The selection of these specific predictor variables was guided by both theoretical relevance and statistical considerations regarding multicollinearity. Initial exploratory analysis revealed strong correlations between several potential predictors. For instance, total expenses and total enrollment showed an extremely high correlation (r > 0.95), as larger boards naturally have higher total expenses. Similarly, various socioeconomic indicators such as median household income, unemployment rates, and low-income status demonstrated substantial overlap in their variation. To address these multicollinearity concerns, we selected representative variables that captured distinct aspects of school board characteristics while minimizing redundancy:

@@

3. Among socioeconomic indicators, the percentage of low-income students and parental education levels were selected as they captured different dimensions of socioeconomic status while maintaining relatively low correlation with each other (r < 0.4).

@@

We evaluate several regression models to understand the relationship between school board characteristics and graduation rates:

### Linear Regression
Our baseline model is a simple linear regression:

\begin{align} 
y_i &= \beta_0 + \beta_1x_{1i} + \beta_2x_{2i} + \beta_3x_{3i} + \beta_4x_{4i} + \beta_5x_{5i} + \epsilon_i
\end{align}

Where $y_i$ is the Four Year Graduation Rate for school board $i$, $\epsilon_i$ is the error term, and $x_{1i}$ to $x_{5i}$ are the predictor variables: percentage of students without a parent holding a post-secondary degree, percentage of low-income students, percentage spent on facilities, expenses per student, and total enrollment.

### Ridge Regression
To address potential multicollinearity between our predictors, we implement Ridge regression:

\begin{align}
\min_{\beta} \sum_{i=1}^{n} (y_i - \beta_0 - \sum_{j=1}^{p} \beta_j x_{ij})^2 + \lambda \sum_{j=1}^{p} \beta_j^2
\end{align}

Where $\lambda$ is the regularization parameter chosen through cross-validation.

### Lasso Regression
We also implement Lasso regression for potential variable selection:

\begin{align}
\min_{\beta} \sum_{i=1}^{n} (y_i - \beta_0 - \sum_{j=1}^{p} \beta_j x_{ij})^2 + \lambda \sum_{j=1}^{p} |\beta_j|
\end{align}

### Polynomial Regression
Our final and chosen model is a polynomial regression of degree 2:

\begin{align}
y_i &= \beta_0 + \sum_{j=1}^{5} (\beta_{j1}x_{ji} + \beta_{j2}x_{ji}^2) + \epsilon_i
\end{align}

Where $\beta_{j1}$ represents the linear term coefficient and $\beta_{j2}$ represents the quadratic term coefficient for each predictor.

@@

We chose the polynomial regression model for several reasons:

1. **Non-linear Relationships**: Initial exploratory data analysis revealed non-linear relationships between several predictors and graduation rates, particularly for expenses per student and total enrollment.

2. **Model Comparison**: Comparing Mean Squared Error (MSE) across models:

[MSE SECTION HERE]

The polynomial regression model showed the lowest MSE, indicating better fit to the data.

@@

### Residual Analysis
```{r}
#| label: fig-residuals
#| fig-cap: Residual Diagnostics for Polynomial Regression Model
#| echo: false
#| message: false
#| warning: false

# Q-Q Plot
ggplot(data.frame(residuals = residuals(poly_model)), aes(sample = residuals)) +
  stat_qq() +
  stat_qq_line() +
  labs(title = "Normal Q-Q Plot",
       x = "Theoretical Quantiles",
       y = "Sample Quantiles") +
  theme_minimal()

@@

@fig-residuals shows our residual diagnostics. The Q-Q plot suggests that residuals are approximately normally distributed, while the Residuals vs Fitted plot shows no clear pattern, indicating homoscedasticity of residuals.

@@

Our polynomial regression analysis reveals several significant relationships between school board characteristics and graduation rates. We present these results in order of statistical significance.

@@

The percentage of households without post-secondary degrees shows the strongest relationship with graduation rates (β = -3.627, p = 0.003). This linear relationship suggests that for every standard deviation increase in the percentage of households without degrees, graduation rates decrease by approximately 3.6 percentage points, holding other factors constant.

@@

The percentage spent on facilities shows a significant negative linear relationship (β = -2.402, p = 0.026), as illustrated in @fig-facilities. This suggests that boards allocating a larger proportion of their budget to facilities tend to have lower graduation rates, possibly indicating trade-offs between infrastructure spending and other educational resources.

@@

Per-student expenses show a significant quadratic relationship (β = 3.444, p = 0.013), as shown in @fig-expenses. The U-shaped relationship suggests that while initial increases in per-student spending may be associated with lower graduation rates, there appears to be a threshold beyond which additional spending becomes positively associated with graduation rates.

@@

### Total Expenses and Enrollment

While neither total expenses nor total enrollment showed statistically significant relationships with graduation rates at the conventional p < 0.05 level, both variables exhibited potentially meaningful trends:

```{r}
#| label: fig-size
#| fig-cap: Relationships Between School Board Size Metrics and Graduation Rates
#| layout-ncol: 2
#| echo: false
#| message: false
#| warning: false
ggplot(data_scaled, aes(x = `Total Expenses`, 
                        y = `Four Year Graduation Rate`)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", 
              formula = y ~ poly(x, 2), 
              se = TRUE) +
  labs(x = "Total Expenses (Standardized)",
       y = "Graduation Rate (Standardized)") +
  theme_minimal()

ggplot(data_scaled, aes(x = `Total Enrolment`, 
                        y = `Four Year Graduation Rate`)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", 
              formula = y ~ poly(x, 2), 
              se = TRUE) +
  labs(x = "Total Enrollment (Standardized)",
       y = "Graduation Rate (Standardized)") +
  theme_minimal()
```

@@

As shown in @fig-size, both total expenses and enrollment suggest potential non-linear relationships with graduation rates, though these relationships did not reach statistical significance in our model.

@@

@fig-fit shows the relationship between predicted and actual graduation rates. The model explains approximately 40.4% of the variance in graduation rates (R² = 0.404), suggesting that while our identified factors are important, there remain other unobserved variables that influence graduation rates.

@@

### Data Constraints
1. Our analysis is cross-sectional, looking at a single academic year (2019-2020)
2. The COVID-19 pandemic may have influenced both spending patterns and graduation rates during this period
3. We lack data on important factors such as:
   - Teacher qualifications and experience
   - Specific program offerings
   - Student-level characteristics

### Methodological Considerations
1. Our polynomial regression model, while capturing non-linear relationships, may not fully represent more complex interactions between variables
2. The R-squared value of 0.404 suggests that substantial variation in graduation rates remains unexplained by our model
3. The standardization of variables, while necessary for our analysis, makes direct interpretation of effect sizes less intuitive

@@

# Appendix {-}

@@

https://data.ontario.ca/dataset/school-board-achievements-and-progress/resource/9d81dc27-32ef-4864-94b5-f09950d00c72

https://data.ontario.ca/dataset/school-board-financial-reports-estimates-revised-estimates-and-financial-statements/resource/56598892-c180-4f0d-bde5-63f971d6ef9a

https://data.ontario.ca/dataset/school-information-and-student-demographics/resource/e0e90bd5-d662-401a-a6d2-60d69ac89d14

>>>>

---
title: "A 1% Increase in Spending on Facilities Decreases Graduation Rates by 0.2%: An Analysis on Ontario School Districts and Related Predictors"
author: Andrew Goh
thanks: "Code and data are available at: https://github.com/mushroomcarbon/schools_and_money"
date: today
date-format: long
abstract: "Using data from Ontario's school board achievement, financial, and demographic datasets, we analyze the relationship between various district-level factors and four-year graduation rates. Through polynomial regression analysis, we find that the percentage of households without post-secondary degrees and relative spending on facilities have significant negative correlations with graduation rates (-0.136, -0.197), while per-student expenses show a non-linear relationship. These findings suggest that socioeconomic factors and resource allocation decisions have statistically significant impacts on academic outcomes, with implications for education policy and resource distribution in Ontario school districts."
format: pdf
header-includes:
  - \usepackage{float}
  - \usepackage{placeins}
number-sections: true
bibliography: references.bib
---

@@

# Load the polynomial model from the RDS file
poly_model <- readRDS(here("models", "poly_model.rds"))
lm_model <- readRDS(here("models", "lm_model.rds"))
ridge_model <- readRDS(here("models", "ridge_model.rds"))
lasso_model <- readRDS(here("models", "lasso_model.rds"))

@@

The relationship between educational resources and academic outcomes has been a central focus of education policy research, particularly in publicly funded school systems. In Ontario, where school boards receive varying levels of funding and serve communities with diverse socioeconomic profiles, understanding how different factors influence student success is crucial for effective policy-making [@campbell].

Recent studies have suggested that both financial resources and community characteristics play significant roles in determining student outcomes [@financial_support_academic_performance; @socioeconomic_status_achievement]. However, the specific mechanisms through which these factors interact, particularly in the Ontario context, remain unclear. This gap in understanding is particularly relevant given Ontario's significant investment in education, with annual spending projected to be approximately \$37.6 billion [@opsba_budget_overview].

Our paper examines how various district-level factors—including financial allocations, socioeconomic indicators, and operational characteristics—correlate with four-year graduation rates across Ontario school boards. We focus specifically on the relationships between graduation rates and five key variables: the percentage of households without post-secondary degrees, the percentage of low-income households, the percentage of the total budget spent on facilities, per-student expenses, and total enrolment.

@@

The results suggest that socioeconomic factors, particularly the percentage of households without post-secondary degrees, show the strongest relationship with graduation rates, with each percentage point increase associated with a 0.136 percentage point decrease in graduation rates. We also find that the percentage of budget spent on facilities has a significant negative relationship with graduation rates, with each additional percentage point allocated to facilities associated with a 0.197 percentage point decrease in graduation rates. While per-student expenses show a U-shaped relationship with graduation rates, this relationship did not reach statistical significance at conventional levels. Our model explains approximately 36.5% of the variance in graduation rates, indicating that while these factors are important, there remain other unobserved variables influencing student outcomes.

@@

## Overview

We use the statistical programming language R [@citer] for the analysis and presentation of the project. The tidyverse [@tidyverse] ecosystem, particularly dplyr [@dplyr], was used for data manipulation. For model development and evaluation, we utilized caret [@caret] for model training and glmnet [@glmnet1; @glmnet2; @glmnet3] for regularized regression methods. The arrow [@arrow] package was employed for efficient data reading and writing, while broom [@broom] was used for converting statistical objects into tidy data frames. Additional file management was facilitated by the here [@here] package.

@@

@tbl-summary presents an overview of the summary statistics of our key variables across Ontario school boards, and @tbl-snapshot1 and @tbl-snapshot2 presents a snapshot of the data itself.

@@

summary_stats %>%
  mutate(
    across(contains("Total Expenses"), ~./1e9),  # Convert to billions
    across(contains("Total Enrolment"), ~./1000)  # Convert to thousands
  ) %>%
  pivot_longer(everything(), 
               names_to = c("Variable", "Statistic"), 
               names_pattern = "(.+)_(.+)") %>%
  pivot_wider(names_from = Statistic, 
             values_from = value) %>%
  gt() %>%
  fmt_number(
    columns = c(Mean, SD, Min, Max, Median),
    decimals = 2
  ) %>%
  cols_label(
    Variable = "Variable",
    Mean = "Mean",
    SD = "Std. Dev.",
    Min = "Minimum",
    Max = "Maximum",
    Median = "Median"
  ) %>%
  tab_footnote(
    footnote = "Total Expenses shown in billions ($), Enrolment in thousands, percentages as raw values, expenses per student in dollars.",
    locations = cells_column_labels(columns = Mean)
  )
```

@@

```{r}
#| label: tbl-snapshot1
#| tbl-cap: "Sample of First Five Rows from Ontario School Board Dataset (Part 1)"
#| echo: false
#| warning: false
#| message: false

# First half of columns
knitr::kable(
  analysis_data |>
    select(`Board Number`,
           `Board Name`,
           Region,
           City,
           `Four Year Graduation Rate`,
           `Total Expenses`) |>
    slice(1:5) |>
    mutate(Region = str_replace(Region, "Region ", "")),
  col.names = c("Board ID",
                "Name",
                "Region",
                "City",
                "Grad Rate",
                "Total Expenses ($)"),
  digits = c(0, 0, 0, 0, 3, 3),
  booktabs = TRUE)
```
```{r}
#| label: tbl-snapshot2
#| tbl-cap: "Sample of First Five Rows from Ontario School Board Dataset (Part 2)"

# Second half of columns
knitr::kable(
  analysis_data |>
    select(`Expenditure on Facilities`,
           `Total Enrolment`,
           percent_low_income,
           percent_no_degree,
           percentage_spent_on_facilities,
           expenses_per_quota) |>
    slice(1:5),
  col.names = c("Facility Expenses ($)",
                "Enrolment",
                "Low Income (%)",
                "No Degree (%)",
                "Facilities (%)",
                "$/Student"),
  digits = c(3, 0, 3, 3, 3, 3),
  booktabs = TRUE)
```

@@

The percentage of students from low-income households ranges from 10.35% to 25.92%, with a mean of 15.77%. This variable describes the socioeconomic composition of each school board's student population, and we can see from @fig-income the diversity of socioeconomic compositions *between* school districts.

@@

The selection of these specific predictor variables was guided by both theoretical relevance and statistical considerations regarding multicollinearity. Initial exploratory analysis revealed strong correlations between several potential predictors. For instance, total expenses and total enrolment showed an extremely high correlation (r \> 0.95), as larger boards naturally have higher total expenses. Similarly, various socioeconomic indicators such as median household income, unemployment rates, and low-income status demonstrated substantial overlap in their variation. To address these multicollinearity concerns, we selected representative variables that captured distinct aspects of school board characteristics while minimizing redundancy:

@@

3.  Among socioeconomic indicators, the percentage of low-income students and parental education levels were selected as they captured different dimensions of socioeconomic status while maintaining relatively low correlation with each other (r \< 0.4).

@@

We evaluate the performance of several regression models to understand the relationship between school board characteristics and graduation rates:

### Linear Regression

We begin with our baseline model, a simple linear regression model:

```{=tex}
\begin{align} 
y_i &= \beta_0 + \beta_1x_{1i} + \beta_2x_{2i} + \beta_3x_{3i} + \beta_4x_{4i} + \beta_5x_{5i} + \epsilon_i
\end{align}
```
Where $y_i$ is the Four Year Graduation Rate for school board $i$, $\epsilon_i$ is the error term, and $x_{1i}$ to $x_{5i}$ are the predictor variables: percentage of students without a parent holding a post-secondary degree, percentage of low-income students, percentage spent on facilities, expenses per student, and total enrolment.

### Shrinkage Regression

To address potential multicollinearity and model overcomplexity between our predictors, we also implemented Ridge and Lasso regression models, respectively:

```{=tex}
\begin{align}
\min_{\beta} \sum_{i=1}^{n} (y_i - \beta_0 - \sum_{j=1}^{p} \beta_j x_{ij})^2 + \lambda \sum_{j=1}^{p} \beta_j^2
\end{align}
```
```{=tex}
\begin{align}
\min_{\beta} \sum_{i=1}^{n} (y_i - \beta_0 - \sum_{j=1}^{p} \beta_j x_{ij})^2 + \lambda \sum_{j=1}^{p} |\beta_j|
\end{align}
```
Where $\lambda$ is the regularization parameter chosen through cross-validation and the x-values are the same as above.

### Polynomial Regression

Finally, a polynomial regression model of degree 2 was fitted to the data to account for non-linear relationships between the data. The model is structured as follows:

```{=tex}
\begin{align}
y_i &= \beta_0 + \sum_{j=1}^{5} (\beta_{j1}x_{ji} + \beta_{j2}x_{ji}^2) + \epsilon_i
\end{align}
```
Where $\beta_{j1}$ represents the linear term coefficient and $\beta_{j2}$ represents the quadratic term coefficient for each predictor.

@@

After comparing the results of the models and assessing how well they each fit to the data respectively, the polynomial regression model was chosen due to multiple reasons:

1.  **Non-linear Relationships**: Initial exploratory data analysis revealed non-linear relationships between several predictors and graduation rates, particularly for expenses per student and total enrolment. This suggests that a polynomial regression model potentially can capture more trends in the data, more accurately, compared to a linear model, though the latter is simpler. The tradeoff in this case is sensible due to the added model complexity actively capturing statistically significant trends in the data.

2.  **Model Comparison**: Comparing Root Mean Squared Error (RMSE) across models:

```{r}
#| echo: false
#| message: false
#| warning: false
#### Workspace setup ####
library(tidyverse)
library(arrow)
library(broom)
library(caret)
library(glmnet)      # For Ridge and Lasso regression

#### Load data ####
analysis_data <- read_parquet(here::here("./data/analysis_data/analysis_data.parquet"))

#### Data Preprocessing ####
data <- analysis_data %>%
  dplyr::select(`Four Year Graduation Rate`, 
                percent_no_degree, 
                percent_low_income, 
                percentage_spent_on_facilities,
                expenses_per_quota,
                `Total Enrolment`)

# Normalize data
data_scaled <- data %>%
  mutate(across(everything(), ~ scale(.) %>% as.vector()))

# Prepare data for glmnet
x <- model.matrix(`Four Year Graduation Rate` ~ ., data_scaled)[, -1]
y <- data_scaled$`Four Year Graduation Rate`

#### Linear Regression Model ####
lm_model <- lm(`Four Year Graduation Rate` ~ ., data = data_scaled)

#### Ridge Regression ####
ridge_cv <- cv.glmnet(x, y, alpha = 0)
ridge_best_lambda <- ridge_cv$lambda.min
ridge_model <- glmnet(x, y, alpha = 0, lambda = ridge_best_lambda)

#### Lasso Regression ####
lasso_cv <- cv.glmnet(x, y, alpha = 1)
lasso_best_lambda <- lasso_cv$lambda.min
lasso_model <- glmnet(x, y, alpha = 1, lambda = lasso_best_lambda)

#### Polynomial Regression ####
poly_model <- lm(`Four Year Graduation Rate` ~ 
                   poly(percent_no_degree, 2) + 
                   poly(percent_low_income, 2) +
                   poly(percentage_spent_on_facilities, 2) +
                   poly(expenses_per_quota, 2) +
                   poly(`Total Enrolment`, 2),
                 data = data_scaled)

#### Model Comparison with MSE ####
# Predictions
lm_preds <- predict(lm_model, newdata = data_scaled)
poly_preds <- predict(poly_model, newdata = data_scaled)
ridge_preds <- predict(ridge_model, newx = x, s = ridge_best_lambda)
lasso_preds <- predict(lasso_model, newx = x, s = lasso_best_lambda)

# Calculate RMSE
lm_rmse <- sqrt(mean((lm_preds - data_scaled$`Four Year Graduation Rate`)^2))
poly_rmse <- sqrt(mean((poly_preds - data_scaled$`Four Year Graduation Rate`)^2))
ridge_rmse <- sqrt(mean((ridge_preds - y)^2))
lasso_rmse <- sqrt(mean((lasso_preds - y)^2))

# Model Comparison Table
model_comparison <- tibble(
  Model = c("Linear Regression", "Polynomial Regression", "Ridge", "Lasso"),
  Metric = c(
    paste("RMSE: ", round(lm_rmse, 3)),
    paste("RMSE: ", round(poly_rmse, 3)),
    paste("RMSE: ", round(ridge_rmse, 3)),
    paste("RMSE: ", round(lasso_rmse, 3))
  )
)

print(model_comparison)
```

The polynomial regression model showed the lowest MSE, which indicates that it has a better fit to the data and is hence a "better" model to use in this case [@validation].

@@

### Residual Analysis

```{r}
#| label: fig-qq
#| fig-cap: Normal QQ Plot for Polynomial Regression Model
#| echo: false
#| message: false
#| warning: false

# Q-Q Plot
ggplot(data.frame(residuals = residuals(poly_model)), aes(sample = residuals)) +
  stat_qq() +
  stat_qq_line() +
  labs(title = "Normal Q-Q Plot",
       x = "Theoretical Quantiles",
       y = "Sample Quantiles") +
  theme_minimal()
```
```{r}
#| label: fig-residuals
#| fig-cap: Residuals vs Fitted Values Plot for Polynomial Regression Model
#| echo: false
#| message: false
#| warning: false

@@

@fig-qq and @fig-residuals show our residual diagnostics. The Q-Q plot suggests that residuals are approximately normally distributed, while the Residuals vs Fitted plot shows no clear pattern, indicating homoscedasticity of residuals.

@@

We present the results from our polynomial regression analysis in order of statistical significance.

@@

The percentage of households without post-secondary degrees shows the strongest relationship with graduation rates (β = -3.627, p = 0.003). After converting from standardized units back to real-world units, this relationship suggests that for every one percentage point increase in households without post-secondary degrees, graduation rates decrease by approximately 0.136 percentage points, holding other factors constant.

@@

The percentage spent on facilities shows a significant negative linear relationship (β = -2.435, p = 0.023), as illustrated in @fig-facilities. After converting from standardized units back to real-world units, this suggests that for each additional percentage point of budget allocated to facilities, graduation rates decrease by about 0.197 percentage points, holding other factors constant. In other words, boards allocating a larger proportion of their budget to facilities tend to have lower graduation rates, possibly indicating trade-offs between infrastructure spending and other educational resources.

@@

Per-student expenses show a marginally significant quadratic relationship (β = 1.932, p = 0.085), as shown in @fig-expenses. After converting from standardized units back to real-world units, while not significant at the conventional p < 0.05 level, there appears to be a U-shaped trend in the relationship between per-student spending and graduation rates. This suggests that graduation rates tend to decrease with initial increases in per-student spending up to a certain threshold, after which additional spending is associated with increasing graduation rates. However, given the lack of statistical significance, these trends should be interpreted with caution.

@@

### Total Expenses and Enrolment

```{r}
#| label: fig-size
#| fig-cap: Relationships Between School Board Size Metrics and Graduation Rates
#| echo: false
#| message: false
#| warning: false
ggplot(data_scaled, aes(x = `Total Enrolment`, 
                        y = `Four Year Graduation Rate`)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", 
              formula = y ~ poly(x, 2), 
              se = TRUE) +
  labs(x = "Total Enrolment (Standardized)",
       y = "Graduation Rate (Standardized)") +
  theme_minimal()
```

@@

As shown in @fig-size, Total Enrolment showed no statistically significant relationship with graduation rates (linear term: β = 0.249, p = 0.842; quadratic term: β = -0.186, p = 0.871). This suggests that the size of the school board, measured by enrollment, has minimal impact on graduation outcomes, with each additional thousand students associated with a non-significant change of 0.00053 percentage points in graduation rates once converted back into real-world units.

@@

@fig-fit shows the relationship between predicted and actual graduation rates. The model explains approximately 36.5% of the variance in graduation rates (R² = 0.3648), suggesting that while our identified factors are important, there remain other unobserved variables that influence graduation rates.

@@

### Data Constraints

1.  Our analysis is cross-sectional, looking at a single academic year (2019-2020)
2.  The COVID-19 pandemic may have influenced both spending patterns and graduation rates during this period
3.  We lack data on important factors such as:
    -   Teacher qualifications and experience
    -   Specific program offerings
    -   Student-level characteristics

### Methodological Considerations

1.  Our polynomial regression model, while capturing non-linear relationships, may not fully represent more complex interactions between variables
2.  The R-squared value of 0.404 suggests that substantial variation in graduation rates remains unexplained by our model
3.  The standardization of variables, while necessary for our analysis, makes direct interpretation of effect sizes less intuitive

@@

# Appendix {.unnumbered}

@@
